---
title: "Data Cleaning"
author: "Rihad Variawa"
date: "11/3/2019"
output: html_document
---

## Executive Summary

The weather dataset is a messy, real-world dataset containing an entire year’s worth of weather data from Boston, USA. Our data cleaning process is composed of three parts:

### First, Eploring Raw Data.

### Second, Tyding Data. Issues addressed in this section include:

* Column headers are values, not variable names: from X1 to X31;
* Meaningless column: X as observational index;
* Variables are stored in both the column: variable measure with max, min, and mean tempretures.

### Third and last, Preparing Data for Analysis, covering:

* day with unecessary leading “X”s
* Date can be united with year, month, and date (yyyy-mm-dd)
* Variable types are not always correct: most characters should be coerced into numerics (for  PrecipitationIn, there are “T” values indicating “Trace”, or 0 in number)
* Finding missing values: in Max.Gust.SpeedMPH
* Dealing with obvious errors: in Max.Humidity and Mean.VisibilityMiles
* Dealing with empty strings: in Events


#### Loading Libraries 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if(!require(tidyr)){install.packages("tidtr")}
if(!require(stringr)){install.packages("stringr")}
if(!require(lubridate)){install.packages("lubridate")}


library(tidyr)
library(stringr)
library(lubridate)
library(dplyr)
```

#### Getting Data

```{r}
# download data
if(!file.exists("weather.rds")){
  download.file("https://assets.datacamp.com/production/repositories/34/datasets/b3c1036d9a60a9dfe0f99051d2474a54f76055ea/weather.rds", "weather.rds")
  dateDownloaded <- date()
}

# load data
weather <- readRDS("weather.rds")
```

#### EDA

```{r}
# verify that weather is a data.frame
class(weather)
```

```{r}
glimpse(weather)
```

```{r}
# preview firat 10 rows
head(weather, n = 10)
```

#### Data Tidying
Issues to address

* Column headers are values, not variable names: from X1 to X31
* Meaningless column: X as observations index
* Variables are stored in both the column: variable measure with max, min, and mean temperature

```{r}
# gather columns
weather2 <- gather(weather, day, value, X1:X31, na.rm = TRUE)
```

```{r}
# preview head
head(weather2)
```

```{r}
# first remove column of row names
weather2 <- weather2[, -1]

# spread the data
weather3 <- spread(weather2, measure, value)

# preview head
head(weather3)
```

#### Preparing Data For Analysis
Issues to address

* day with unnecssary leading "X"s
* Date can be united with year, month, and date(yyyy-mm-dd)

```{r}
# remove X from day column
weather3$day <- str_replace(weather3$day, "X", "")

# unite year, month and day columns
weather4 <- unite(weather3, date, year, month, day, sep = "-")

# convert date column to proper date format using lubridates's ymd()
weather4$date <- ymd(weather4$date)

# rearrange columns using dplyr select()
weather5 <- weather4 %>% 
  select(date, Events, CloudCover:WindDirDegrees)

# preview weather5
head(weather5)
```

Additional issues include:

* Variable types are not always correct: most characters should be coerced into numerics (for  PrecipitationIn, there are “T” values indicating “Trace”, or 0 in number)

```{r}
# preview structure of weather5
str(weather5)
```

```{r}
# replace "T" with "0" (T = trace)
weather5$PrecipitationIn <- str_replace(weather5$Precipitation, "T", "0")

# convert characters to numeric
weather6 <- mutate_at(weather5, vars(CloudCover:WindDirDegrees), funs(as.numeric))

# preview result
str(weather6)
```

Further issues include:

* Finding missing values: in Max.Gust.SpeedMPH
* Dealing with obvious errors: in Max.Humidity and Mean.VisibilityMiles
* Dealing with empty strings: in Events

```{r}
# find missing values
summary(weather6)
```

```{r}
# preview full rows for observations missing values Max.Gust.SpeedMPH
weather6[is.na(weather6$Max.Gust.SpeedMPH), ]
```

```{r}
# preview full row for record with Max.Humidity of 1000 which is an obvious error: humidity from 0 to 100, and use 100 instear
weather6$Max.Humidity[weather6$Max.Humidity == 1000] <- 100

# preview full row with -1 value for Mean.VisibilityMiles, which should be above 0, and use 10 instead
weather6$Mean.VisibilityMiles[weather6$Mean.VisibilityMiles == -1] <- 10

# replace empty cells in events column
weather6$Events[weather6$Events == ""] <- "None"
```

